---
title: "Basic Syntax and Operators"
output: 
  learnr::tutorial:
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
set.seed(1)
random_numbers <- rf(1000, 10, 100)
```

## Calculating Mean and Standard Deviation

### Calculating the Mean
We can run the following code to create a vector of 1000 random numbers. The function `set.seed()` ensures that the process used to generate random numbers is the same across computers.

```{r random-numbers-setup, echo=TRUE, eval=TRUE}
set.seed(1)
random_numbers <- rf(1000, 10, 100)
```

Write a function that gives you the sum of `random_numbers` and saves it to a new variable called `numbers_sum`:

```{r sum, exercise=TRUE, exercise.setup}

```

Write a function that gives you the number of items in the `random_numbers` vector and saves it to a new variable called `numbers_count`:
```{r count, exercise=TRUE}

```

<div id="count-hint">
**Hint:** To count the number of items in a vector, use the `length()` function.
</div>

Now write a function that uses the above two functions to calculate the average of `random_numbers` and assign it to a new variable called `this_mean`. 

```{r mean, exercise=TRUE}

```

```{r question-mean}
question("What number do you get for the mean of `random_numbers`?",
  answer("1000"),
  answer("1018"),
  answer("1.018", correct = TRUE),
  answer("0.982"),
  allow_retry = TRUE
)
```

R actually has a built in function to calculate mean for you, so you don't have to remember how to build it from scratch each time! Check your above answer by using the `mean()` function on the `random_numbers` vector.
```{r mean2, exercise=TRUE}

```

### Calculating the Standard Deviation

The formula for calculating standard deviation is as follows:

$s^2 = \sqrt{\sum_{i=1}^n (y_i - \bar{y})^2} / n-1$

Can you convert this into code? 

We'll introduce a new concept, *vectorization*. This means that an operation like subtracting will act on all numbers in a vector at the same time.

Subtract `this_mean` from the `random_numbers` vector. Did each number in `random_numbers` change?
```{r sd-vector, exercise=TRUE}

```

[Insert more content here]

```{r question-sd}
question("What number do you get for the standard deviation of `random_numbers`?",
  answer("0.489704", correct = TRUE),
  answer("1018"),
  answer("1.018"),
  answer("0.982"),
  allow_retry = TRUE
)
```

Check if you calculated the standard deviation correctly by using the `sd()` function on the `random_numbers` vector.
```{r sd2, exercise=TRUE}

```

### Making a Histogram of Our Numbers

What do these random numbers look like, anyway? We can use an R package called ggplot2 to visualize the distribution of our random numbers.

Run the following code to visualize the original distribution of `random_numbers`.
```{r viz, exercise=TRUE}
hist(random_numbers)
```

Notice how most of the values are concentrated on the left-hand side of the graph, while there is a longer "tail" to the right? Counterintuitively, this is known as a right-skewed distribution. When we see a distribution like this, one common thing to do is to normalize it.

This is also known as *calculating a z-score*, which we will cover next.

## Calculating a Z-Score

```{r z-scores-setup}
set.seed(1)
random_numbers <- rf(1000, 10, 100)
this_mean <- mean(random_numbers)
this_sd <- sd(random_numbers)
```

The formula for calculating a z-score is as follows:

$x_i - x_bar / sd$

This can be calculated for each value in `random_numbers` in context of the larger set of values.

Can you translate this formula into code?

Assume `random_numbers`, `this_mean`, and `this_sd` are in your environment. Write a formula to transform `random_numbers` into z-scores, and assign it to the new variable `normalized_data`.
```{r norm-data, exercise=TRUE, exercise.setup = "z-scores-setup"}

```

```{r norm-setup, exercise.setup = "z-scores-setup"}
normalized_data <- (random_numbers - this_mean) / this_sd
```

Take the mean of `normalized_data` and assign it to a variable called `normalized_mean`.
```{r norm-mean, exercise=TRUE, exercise.setup = "norm-setup"}

```

Take the standard deviation of `normalized_data` and assign it to a variable called `normalized_sd`.
```{r norm-sd, exercise=TRUE, exercise.setup = "norm-setup"}

```


```{r question-normalized}
question("What is the value of `normalized_mean`? What is the value of `normalized_sd`?",
  answer("`normalized_mean` is 0, and `normalized_sd` is 1", correct = TRUE),
  answer("`normalized_mean` is 0, and `normalized_sd` is 0.489704"),
  answer("`normalized_mean` is 1.018126, and `normalized_sd` is 1"),
  answer("`normalized_mean` is 1.018126, and `normalized_sd` is 0.489704"),
  allow_retry = TRUE
)
```

### Making a Histogram of Z-scores

Let's plot the Z-scores and see if our values are still skewed!

```{r viz2, exercise=TRUE, , exercise.setup = "norm-setup"}
hist(normalized_data)
```

```{r question-skewed}
question("Is this data skewed?",
  answer("Yes, it is right-skewed", correct = TRUE),
  answer("Yes, it is left-skewed"),
  answer("No, this data is not skewed"),
  allow_retry = TRUE
)
```

## Calculating a T-Score

T-tests are used to determine if two sample means are equal. The formula for calculating a t-score is as follows: 

$(mu_1 - mu_2) / sqrt(s_1/n_1 + s_2/n_2)$

where $mu_i$ is the mean of the first or second set of data, $s_i$ is the sample standard deviation of the ith set of data, and $n_i$ is the sample size of the $i$th set of data.

We'll first create two data sets of random numbers following a normal distribution:

```{r echo=TRUE, eval=TRUE}
data_1 <- rnorm(1000, 3)
data_2 <- rnorm(100, 2)
```

Here's how we'll calculate the values for the first data set:

```{r echo=TRUE, eval=TRUE}
mu_1 <- mean(data_1)
s_1 <- sd(data_1)^2
n_1 <- length(data_1)
```

Can you do the same for `data_2`?

```{r data2, exercise=TRUE}
```

```{r t-score-setup}
mu_2 <- mean(data_2)
s_2 <- sd(data_2)^2
n_2 <- length(data_2)
```

Now, can you translate the t-score formula into code?

```{r t_score, exercise=TRUE, exercise.setup = "t-score-setup"}

```

```{r question-tscore}
question("What did you get for the t-score?",
  answer("77.897"),
  answer("8.998", correct = TRUE),
  answer("32.662"),
  answer("-13.540"),
  allow_retry = TRUE
)
```

The t-score's meaning depends on your sample sizes, but in general t-scores close to 0 imply that the means are not statistically distinguishable and large t-scores (e.g. t > 3) imply the data have different means.

## Performing a T-Test

Once again, R has a built in function that will perform a T-test for us, aptly named `t.test()`. Look up the arguments the function `t.test()` takes, and perform a T-test on `data_1` and `data_2`.

```{r t_test, exercise=TRUE, exercise.setup = "t-score-setup"}

```

```{r question-ttest}
question("What are the sample means, and are they distinguishable from each other?",
  answer("The mean of `data_1` is 1.904, the mean of `data_2` is 2.943, and they are not distinguishable from each other."),
  answer("The mean of `data_1` is 1.904, the mean of `data_2` is 2.943, and they are distinguishable from each other."),
  answer("The mean of `data_1` is 2.943, the mean of `data_2` is 1.904, and they are not distinguishable from each other."),
  answer("The mean of `data_1` is 2.943, the mean of `data_2` is 1.904, and they are distinguishable from each other.", correct = TRUE),
  allow_retry = TRUE
)
```
